---
title: "Reproducible Research: Peer Assessment 1"
author: "Juan Villarreal"
date: "March 25, 2017"
output: html_document 
html_document: PA1_template.html
keep_md: true
---
## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "asis")
```

## 1 Loading and preprocessing the data
```{r LoadData1}
dataOriginal <- read.csv(unzip("activity.zip", "activity.csv"), header = TRUE, sep=",")
## ignoring NA Steps values in a new dataset
dataNoNA <- na.omit(dataOriginal)
```

## 2 Histogram of the total number of steps taken each day
```{r barPlo2}
## TSDAy = Total Steps by Day
TSDay <- aggregate(steps ~ date, dataNoNA, sum)
par(mar=c(10,5,1,1))
#png('F://R WorkingDirectory//Reproducible_research1//Figures//Figure1.png')
barplot(height = TSDay$steps, width = c(5),names.arg = TSDay$date, xlab = "", ylab = "# of Steps", main = "Total steps by day", col = "blue", las=2)
#dev.off()
```

## 3 Mean and median number of steps taken each day
```{r Table3}
library(xtable)
MeanTSDay <- aggregate(steps ~ date, TSDay, mean)
MedianTSDay <- aggregate(steps ~ date, TSDay, median)
MediaMeanTSDay <- xtable(merge(x=MeanTSDay,y=MedianTSDay, by = "date", all = TRUE))
names(MediaMeanTSDay) <- c("Date", "Mean","Median")
print(MediaMeanTSDay, type= "html")
```

## 4 Time series plot of the average number of steps taken?
```{r Timeline4}
#3ASInterval = Average Steps by Interval
ASInterval <- aggregate(steps ~ interval, dataNoNA, mean)
#png('F://R WorkingDirectory//Reproducible_research1//Figures//Figure2.png')
plot(ASInterval$interval, ASInterval$steps, type = "l", xlab = "Interval Number", ylab = "Average Steps", main = "Average Steps by Interval", sub = "between 2012-10-02 and 2012-11-29")
#dev.off()
```

## 5 The 5-minute interval that, on average, contains the maximum number of steps?
```{r MaximumInterval5}
FiveminMax <- xtable(ASInterval[ASInterval$steps==max(ASInterval$steps),])
print(FiveminMax, type= "html")
```

## 6 Imputing missing values, Merging the mean for each interval to fill the steps NA values with realistics values
```{r Imputing6}
 library(plyr)
## Selecting only NA rows for steps column
dataNA <- dataOriginal[is.na(dataOriginal$steps),] 
## Merging the mean for each interval to fill the steps NA values with something realistic
dataMerged <- subset(merge(x=dataNA, y=ASInterval, by = "interval", all.x = TRUE), select = c(4,3,1))
## Giving right names to the merged data set to union with the dataset that have all values except by rows with Null steps values
colnames(dataMerged) <- c("steps","date","interval")
dataImputed <- rbind.fill(dataMerged, dataNoNA)
```

## 7 Histogram of the total number of steps taken each day after missing values are imputed 
```{r Imputing7}
TSDayImputed <- aggregate(steps ~ date, dataImputed, sum)
par(mar=c(10,5,1,1))
#png('F://R WorkingDirectory//Reproducible_research1//Figures//Figure3.png')
barplot(height = TSDayImputed$steps, width = c(5),names.arg = TSDayImputed$date, xlab = "", ylab = "# of Steps", main = "Total steps by day With imputed data", col = "blue", las=2)
#dev.off()
```

```{r Imputing8}
## Performing same calculations on step 3 but with the imputed dataset 
MeanTSDayImputed <- aggregate(steps ~ date, TSDayImputed, mean)
MedianTSDayImputed <- aggregate(steps ~ date, TSDayImputed, median)
MediaMeanTSDayImputed <- xtable(merge(x=MeanTSDayImputed,y=MedianTSDayImputed, by = "date", all = TRUE))
names(MediaMeanTSDayImputed) <- c("Date", "Mean","Median")
print(MediaMeanTSDayImputed, type= "html")

## Performing same calculations on step 4 but with the imputed dataset 
ASIntervalImputed <- aggregate(steps ~ interval, dataImputed, mean)
#png('F://R WorkingDirectory//Reproducible_research1//Figures//Figure4.png')
plot(ASIntervalImputed$interval, ASIntervalImputed$steps, type = "l", xlab = "Interval Number", ylab = "Average Steps", main = "Average Steps by Interval", sub = "between 2012-10-01 and 2012-11-29 Including Imputed data")
#dev.off()

## Performing same calculations on step 5 but with the imputed dataset 
FiveminMaxImputed <- xtable(ASIntervalImputed[ASIntervalImputed$steps==max(ASIntervalImputed$steps),])
print(FiveminMaxImputed, type= "html")
```

## There isn't a big difference if we ignore missing NA values in terms of our specific analysis, The avarege by day for the missing values can be found in the histogram chart, in the list, the maximum vaule with steps still the same than ignoring the NA values.


## 8 Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

## Are there differences in activity patterns between weekdays and weekends?
```{r Daytypeanalysis8}
## Adding the column to flag the rows as Weekdays by default all are weekedays
dataImputed <- cbind(Daytype = 'Weekday', dataImputed)
## Converting date text column into a Date column
dataImputed$date <- as.Date(dataImputed$date)
## Updating the rows in the column Daytype in case they are weekends only
dataImputed$Daytype  <- ifelse(weekdays(dataImputed$date) %in% c("Saturday", "Sunday"), "Weekend", "weekday")
## Generating the mean for the steps by interval and DayType
ASImputedDaytype <- aggregate(steps ~ interval + Daytype, dataImputed, mean)
library(ggplot2)
par(mar=c(1,1,1,1))
#png('F://R WorkingDirectory//Reproducible_research1//Figures//Figure5.png')
qplot(x=interval, y=steps, data = ASImputedDaytype, facets = . ~ Daytype, main = "Benchmark between Weekend and Weekdays average steps done by interval")
#dev.off()
```

## There is a clear difference in avarage steps done during weekend days and weekday days for same intervals
